Anomalous waves triggered by abrupt depth changes: laboratory experiments and truncated KdV statistical mechanics

### 2020-04-23 paper2_JNLS/paper.tex
NM

\begin{abstract} 
Recent laboratory experiments of \boetal demonstrate that an abrupt change in bottom topography can trigger anomalous statistics in randomized surface waves. Motivated by these observations, \maetal developed a theoretical framework, based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) system, that successfully captures key qualitative features of the experiments, including the robust emergence of anomalous statistics and heightened skewness in the outgoing wavefield. Here, we extend these parallel experimental and modeling efforts with several new findings that have resulted from a synergetic interaction between the two. By precisely relating model parameters to physical ones, we calibrate the model inverse temperature to the specific conditions present in the experiments, thereby permitting a quantitative comparison. We find theoretically predicted distributions of surface displacement to match the experimental measurements with surprising detail. Prompted by the presence of surface slope in the TKdV Hamiltonian, we present new experimental measurements on surface slope statistics and compare them to model predictions. Analysis of some deterministic trajectories of TKdV elucidates the experimental length and time scales required for the statistical transition to a skewed state. Finally, the theory predicts a peculiar relationship between the outgoing displacement skewness and the change in slope variance, specifically how their ratio depends on the wave amplitude and depth ratio. New experimental measurements confirm this prediction in spectacular fashion.
\end{abstract}
%New experimental measurements provide convincing evidence for this prediction.

@@

	Figure \ref{ExpSpatialStats} shows how these statistics vary in the vicinity of the ADC, located at $x = 0$, for both displacement and slope. First, the standard deviation of displacement, $\etastd$, gives the coarsest possible estimate for the amplitude of waves. Figure \ref{ExpSpatialStats}(a) shows that while $\etastd$ increases with driving amplitude, it remains relatively uniform in space for each individual experiment, indicating that the overall amplitude of the wave train is not significantly altered by the presence of the ADC. The skewness and kurtosis, however, respond strongly to the ADC as long as the amplitude is sufficiently high. As seen in \ref{ExpSpatialStats}(b)--(c), both the skewness and kurtosis are relatively small upstream of the ADC, indicating nearly Gaussian statistics, but then increase dramatically downstream and reach a peak near $x = 15$ cm. Interestingly, the location of the peak is the same for skewness and kurtosis and appears insensitive to driving amplitude. The maximum values of skewness and kurtosis (roughly 0.9 and 0.7 respectively) seen in the figure indicate a significant departure from Gaussian statistics, as is constant with the histogram in Fig.~\ref{ExpDiagStats}(e). \boetal found that, once a threshold driving amplitude is exceeded (roughly $\Dphi = 0.5^{\circ}$), the displacement statistics in this anomalous region are robustly described by the gamma distribution across all of the experiments. We note that we have gathered data for 15 different driving amplitudes, but only display 6 in Fig.~\ref{ExpSpatialStats} to avoid clutter.

@@

% Remarkable, spectacular, splendid, stupendous, striking, amazing, exquisite
% Compelling, convincing

@@

	Finally, Fig.~\ref{trajectories} sheds some light on the timescales required for wave evolution in the TKdV framework. It appears that the normalized time of $t=10$ is roughly the correct timescale to observe substantial wave dynamics. More precisely, both the left-going waves in Fig.~\ref{trajectories}(a) and the right-going waves in Fig.~\ref{trajectories}(b) require a dimensionless time of about $t=7$ to traverse the entire periodic domain. For the representative experiments, the characteristic timescale, $\timescale$, is 2.8~sec upstream and 5.7~sec downstream. Hence, a dimensionless time of $t=7$ corresponds to roughly 20~sec and 40~sec upstream and downstream respectively. Given the wave speeds ($c = $ 110 cm/s upstream and 54 cm/s downstream), these timescales can be converted to distances traveled by the waves. In fact, a simple calculation using the definitions in Section \ref{theory} gives the distance traveled as $7 \lamfac g/(2 \pi \freqp^2)$, which, due to a serendipitous cancellation, is independent of depth. Hence, the distance traveled by the waves over a dimensionless time of $t=7$ is the same value upstream and downstream,  roughly 22 meters. Note that this distance is significantly greater than the 6-meter length of the experimental wave tank.

@@

Inserting \eqref{H3convert} and \eqref{H2convert} into \eqref{H3H2ratio}, and using the definitions of $C_2$ and $C_3$ from \eqref{C3C2}, gives the remarkably explicit formula
\begin{equation}
\label{SkewPred}
\frac{\skwdn(\eta)} {\vardn(\eta_x) - \varup(\eta_x)} = \frac{1}{3} \epsup^{-3} \dratdn^3 
%= \frac{1}{3} \left( \frac{\ddn}{\etastd} \right)^3
\end{equation}
This formula links the skewness of the outgoing wave field to the change in the variance of surface slope. In particular, it predicts their ratio to scale as the {\em inverse cube of wave amplitude}, $\epsup^{-3}$, and the {\em cube of the depth ratio}, $\dratdn^3$. It is a rather unexpected relationship, as it would have been difficult to anticipate that the displacement skewness and slope variance, among all possible variable combinations, are so intimately related.

@@

	With these locations selected, we evaluate the ratio $\skw(\eta) / \Delta \var(\eta_x)$ for each experiment and plot the result against dimensionless wave amplitude, $\epsup = \etastd / \dup$ on a log-log scale in Fig.~\ref{SkewRat}(c). While the data shows significant variation at small amplitudes, it shows a well defined decreasing trend at larger amplitudes. In particular, the red dashed line shows the power-law $\epsup^{-3}$ predicted by \eqref{SkewPred}, which captures the experimental trend remarkably well.

	\new{We remark that Eq.~\eqref{SkewPred} bears similarity to a formula derived by Onorato \& Suret \cite{onorato2016twenty}, in the context of {\em constant-depth} KdV, that relates variations in skewness to variations in spectral bandwidth. Indeed, since $\eta_t \approx c \eta_x$ to leading order, the term $\var(\eta_x)$ in Eq.~\eqref{SkewPred} could be (approximately) interpreted as the spectral bandwidth.
We note that an analogous formula has been derived for the nonlinear Schr{\"o}dinger equation that relates kurtosis to bandwidth variations \cite{onorato2016origin}. The present study differs in that it predicts a moment-bandwidth relation for waves propagating over {\em variable} depth, where it is the depth change that creates system disequilibrium. A small transition region connects two near-equilibrium systems, i.e. one far upstream and one far downstream. The link between these two equilibrium states is made mathematically precise by the statistical matching condition \eqref{statmatch}, which is the key ingredient needed to derive Eq.~\eqref{SkewPred}. This is in contrast to the simpler situation of waves propagating over constant depth, where spectral variations and the associated moment variations arise from out-of-equilibrium initial conditions \cite{onorato2016twenty, onorato2016origin}.}

>>>>

\begin{abstract} 
Recent laboratory experiments of \boetal demonstrate that an abrupt change in bottom topography can trigger anomalous statistics in randomized surface waves. Motivated by these observations, \maetal developed a theoretical framework, based on deterministic and statistical analysis of the truncated Korteweg–de Vries (TKdV) system, that successfully captures key qualitative features of the experiments, including the robust emergence of anomalous statistics and heightened skewness in the outgoing wavefield. Here, we extend these parallel experimental and modeling efforts with several new findings that have resulted from a synergetic interaction between the two. By precisely relating model parameters to physical ones, we calibrate the model inverse temperature to the specific conditions present in the experiments, thereby permitting a quantitative comparison. We find theoretically predicted distributions of surface displacement to match the experimental measurements with surprising detail. Prompted by the presence of surface slope in the TKdV Hamiltonian, we present new experimental measurements on surface slope statistics and compare them to model predictions. Analysis of some deterministic trajectories of TKdV elucidates the experimental length and time scales required for the statistical transition to a skewed state. Finally, the theory predicts a peculiar relationship between the outgoing displacement skewness and the change in slope variance, specifically how their ratio depends on the wave amplitude and depth ratio. \new{New experimental measurements provide convincing evidence for this prediction.}
\end{abstract}

% ORIGINAL: New experimental measurements confirm this prediction in spectacular fashion.

% OTHER
% New experimental measurements provide convincing evidence for this prediction.
% convincing, conclusive, compelling.

@@

	Figure \ref{ExpSpatialStats} shows how these statistics vary in the vicinity of the ADC, located at $x = 0$, for both displacement and slope. First, the standard deviation of displacement, $\etastd$, gives the coarsest possible estimate for the amplitude of waves. Figure \ref{ExpSpatialStats}(a) shows that while $\etastd$ increases with driving amplitude, it remains relatively uniform in space for each individual experiment, indicating that the overall amplitude of the wave train is not significantly altered by the presence of the ADC. The skewness and kurtosis, however, respond strongly to the ADC as long as the amplitude is sufficiently high. As seen in \ref{ExpSpatialStats}(b)--(c), both the skewness and kurtosis are relatively small upstream of the ADC, indicating nearly Gaussian statistics, but then increase dramatically downstream and reach a peak near $x = 15$ cm. Interestingly, the location of the peak is the same for skewness and kurtosis and \new{is} insensitive to driving amplitude. The maximum values of skewness and kurtosis (approximately 0.9 and 0.7 respectively) seen in the figure indicate a significant departure from Gaussian statistics, as is constant with the histogram in Fig.~\ref{ExpDiagStats}(e). \boetal found that, once a threshold driving amplitude is exceeded (roughly $\Dphi = 0.5^{\circ}$), the displacement statistics in this anomalous region are robustly described by the gamma distribution across all of the experiments. We note that we have gathered data for 15 different driving amplitudes, but only display 6 of those in Fig.~\ref{ExpSpatialStats} to avoid clutter.

@@

@@

	Finally, Fig.~\ref{trajectories} sheds light on the timescales required for wave evolution in the TKdV framework. It appears that the normalized time of $t=10$ is \new{approximately} the correct timescale to observe substantial wave dynamics. More precisely, both the left-going waves in Fig.~\ref{trajectories}(a) and the right-going waves in Fig.~\ref{trajectories}(b) require a dimensionless time of about $t=7$ to traverse the entire periodic domain. For the representative experiments, the characteristic timescale, $\timescale$, is 2.8~sec upstream and 5.7~sec downstream. Hence, a dimensionless time of $t=7$ corresponds to 20~sec and 40~sec upstream and downstream respectively. Given the wave speeds ($c = $ 110 cm/s upstream and 54 cm/s downstream), these timescales can be converted to distances traveled by the waves. In fact, a simple calculation using the definitions in Section \ref{theory} gives the distance traveled as $7 \lamfac g/(2 \pi \freqp^2)$, which, due to a serendipitous cancellation, is independent of depth. Hence, the distance traveled by the waves over a dimensionless time of $t=7$ is the same value upstream and downstream, 22 meters. Note that this distance is significantly greater than the 6-meter length of the experimental wave tank.

@@

Inserting \eqref{H3convert} and \eqref{H2convert} into \eqref{H3H2ratio}, and using the definitions of $C_2$ and $C_3$ from \eqref{C3C2}, gives the explicit formula
\begin{equation}
\label{SkewPred}
\frac{\skwdn(\eta)} {\vardn(\eta_x) - \varup(\eta_x)} = \frac{1}{3} \epsup^{-3} \dratdn^3 
%= \frac{1}{3} \left( \frac{\ddn}{\etastd} \right)^3
\end{equation}
This formula links the skewness of the outgoing wave field to the change in the variance of surface slope. In particular, it predicts their ratio to scale as the {\em inverse cube of wave amplitude}, $\epsup^{-3}$, and the {\em cube of the depth ratio}, $\dratdn^3$. It is a rather unexpected relationship, as it would have been difficult to anticipate that the displacement skewness and slope variance, among all possible variable combinations, are so intimately related.

@@

	With these locations selected, we evaluate the ratio $\skw(\eta) / \Delta \var(\eta_x)$ for each experiment and plot the result against dimensionless wave amplitude, $\epsup = \etastd / \dup$ on a log-log scale in Fig.~\ref{SkewRat}(c). While the data shows significant variation at small amplitudes, it shows a well defined decreasing trend at larger amplitudes. In particular, the red dashed line shows the power-law $\epsup^{-3}$ predicted by \eqref{SkewPred}, which captures the experimental trend \new{closely}.

	\new{We remark that Eq.~\eqref{SkewPred} bears similarity to a formula derived by Onorato \& Suret \cite{onorato2016twenty}, in the context of {\em constant-depth} KdV, that relates variations in skewness to variations in spectral bandwidth. Indeed, since $\eta_t \approx c \eta_x$ to leading order, the term $\var(\eta_x)$ in Eq.~\eqref{SkewPred} could be (approximately) interpreted as the spectral bandwidth.
We note that an analogous formula has been derived for the nonlinear Schr{\"o}dinger equation relating kurtosis to bandwidth variations \cite{onorato2016origin}. The present study differs in that it predicts a moment-bandwidth relation for waves propagating over {\em variable} depth, where it is the depth change that creates system disequilibrium. A small transition region connects two near-equilibrium systems, i.e. one far upstream and one far downstream. The link between these two equilibrium states is made mathematically precise by the statistical matching condition \eqref{statmatch}, which is the key ingredient needed to derive Eq.~\eqref{SkewPred}. This is in contrast to the simpler situation of waves propagating over constant depth, where spectral variations and the associated moment variations arise from out-of-equilibrium initial conditions \cite{onorato2016twenty, onorato2016origin}.}
